\chapter{Implementation of \enquote{lox}}

This chapter discusses the major design decisions and notable implementation details behind \code{lox}.
The last section contains a short overview over all features offered by the library. As already mentioned in the introduction, \code{lox} is licensed as MIT/Apache-2.0 and can be found here: \url{https://github.com/LukasKalbertodt/lox}.

% TODO: one crate, many Cargo features
The library consists of a single crate (one minor exception is explained later) which is very common for small to medium-sized Rust libraries.
As \code{lox} contains a variety of features, it is likely that some users do not need all of them.
To avoid compiling parts of the library that are unused, \code{lox} uses \emph{Cargo features} to let the user disable those parts individually.
For example, if a user does not read from or write to files, the IO module can be disabled completely and thus will not be compiled.

To maintain a high code quality, various automated tests and checks are used.
These are further explained in the respective sections in this chapter.
All of these tests are automatically executed on a \emph{continuous integration} (CI) server whenever commits are pushed to the repository or a \emph{pull request} is created.
The CI also builds the documentation and makes sure that no compiler warnings are produced.

% TODO: maybe put this in a box or make it stand out a bit?
As an aside:
the term \enquote{lox} is often used as an abbreviation for \emph{\textbf{l}iquid \textbf{ox}ygen}, a pale-blue liquid usually obtained by cooling elemental oxygen below its boiling point of 90.188\,K.
This substance is used in many areas, playing a particularly important role in the aerospace industry where it is commonly used as a cryogenic oxidizer in rockets.
Since rust (as a chemical compound\footnote{The programming language \emph{Rust} is not even named after the chemical compound, but after the \enquote{Rust} family of fungi. Source: \url{https://www.reddit.com/r/rust/comments/27jvdt/}}) is a product of an oxidation reaction and rockets are usually fast, using \enquote{lox} as name for this library seemed fitting.
However, the name was mainly chosen because it is short and pronounceable.

% TODO: lox logo?

\newpage
\input{chapters/4-1-basic-layout-considerations}

\newpage
\input{chapters/4-2-mesh-traits}

\newpage
\input{chapters/4-3-io}

\newpage
\input{chapters/4-4-performance-optimizations}

\newpage
\input{chapters/4-5-summary}
